---
title: "3. Finch Data Example"
---

In Part 3 of this lab, you will be introduced to R scripts. You will use the Finches dataset to learn how to read a tabular dataset into R, inspect it, and perform basic statistical summaries on it. You will estimate means, standard deviations, standard errors of the mean, and confidence intervals around the mean. You will also learn how to plot histograms and barplots.

### Download the data

Download the data and save it in the project directory (folder) you created earlier. For example, if you created your project in the **Documents** folder and named it **Org Bio Data Lab 1** then save the data into the **Org Bio Data Lab 1** folder.

You will know you saved it to the correct location if the Excel file shows up in the Files panel in RStudio next to your R Project file.

<a href="finches_data.xlsx" class="btn btn-default">Download</a>

### Reading data into R

This requires installing some packages in R, which let you use functions not found in the standard R installation.

To download package, click the **Package** tab to the right and click **Install**. Type the name of the package, `tidyverse`, and click **Install**. You only have to do this once on each computer you are using.

Now you're ready to load the packages.

```{r}
library(readxl)                # load readxl, for reading Excel files
library(tidyverse)             # load tidyverse, for working with datasets
```

You will see a warning and some messages when you run this code, but those can be safely ignored for now. 

Read the data.

```{r}
finches <- read_excel("finches_data.xlsx")
finches
```

Explain what they are seeing

Show ways to view the data:

```{r}
glimpse(finches)
```

### Plotting a histogram

Plotting is most easily accomplished using the `ggplot()` function and its associated helper functions.

For example, to plot a histogram of the `beak_length` variable, use the following code. In this example, you will set the x axis to `beak_length` and setting the number of bins to use for the histogram to `12`.

```{r histogram}
ggplot(finches, aes(x = beak_length)) +
  geom_histogram(bins = 12)
```
You can plot histograms of other columns by changing the `x = ` part of the code. 

You can also experiment with different numbers of bins. The goal is to have enough that you can see a hump-shaped distribution, but not so many that you have too many bins with few or no observations. Try plotting a histogram with 50 bins, and another with 3 bins. Not very informational, are they?

Sometimes we want to plot separate histograms for differet subsets of the data. One easy way to do this is with the `facet_wrap()` function in ggplot. Note that you have to add a `+` operator after the `geom_histogram()` line.

```{r grouped-histogram}
ggplot(
  data = finches, 
  mapping = aes(x = beak_length, fill = outcome)
) +
  geom_histogram(bins = 14) +
  facet_wrap(~ outcome, ncol = 1)
```

The last thing you might want to do is spruce up your plot with a title and better axis labels. This is done with the `labs()` function:

```{r grouped-histogram-labels}
ggplot(
  data = finches, 
  mapping = aes(x = beak_length, fill = outcome)
) +
  geom_histogram(bins = 14) +
  facet_wrap(~ outcome, ncol = 1) +
  labs(
    x = "Beak Length (mm)",
    y = "Number of Birds",
    title = "Beak length vs. drought survival"
  )
```

#### Export your figure

To use your figure in a Microsoft Word document, PowerPoint presentation, or other software, you first need to save the image. The most reproducible way to do so is to use the `ggsave()` function.

By default, `ggsave()` will save the last plot displayed, so make sure the plot you want to save is displayed in the **Plots** tab in the bottom right panel before running this code.


```{r save-histogram, eval=FALSE}
ggsave("Beak Length Histogram.png",     # name your file, use .png
       width = 6.5, height = 6.5,       # height and width of output
       units = "in")                    # what units are height and width? in=inches
```

### Summarizing datasets

Scientists often want a quick summary of a dataset. The `summary()` function provides just that, and is especially useful for seeing summary statistics of numerical variables. Running this function prints a summary to the console, with the minimum value, median value, mean, maximum value, and first and third quartiles given for each numerical variable:

```{r}
summary(finches)
```

If we want to do anything more than, then we need to use functions provided in the tidyverse to summarize the data. The following code will summarize a dataset by calculating the mean, standard deviation, standard error of the mean (SEM), and 95% confidence intervals of the `beak_length` variable.

```{r}
summary <- finches %>% 
  summarize(mean = mean(beak_length),
            sd = sd(beak_length),
            n = n()) %>% 
  mutate(sem = sd / sqrt(n),
         upper = mean + 2 * sem,
         lower = mean - 2 * sem)
summary
```

If we wanted to get separate summaries for different subsets of the data, we can add the `group_by()` function like this:

```{r}
grouped_summary <- finches %>% 
  group_by(outcome) %>% 
  summarize(mean = mean(beak_length),
            sd = sd(beak_length),
            n = n()) %>% 
  mutate(sem = sd / sqrt(n),
         upper = mean + 2 * sem,
         lower = mean - 2 * sem)
grouped_summary
```

To plot these means and confidence intervals:

```{r column-chart, fig.width=4, fig.height=4}
ggplot(
  data = grouped_summary, 
  mapping = aes(x = outcome, y = mean, fill = outcome)
) +
  geom_col() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = .3) +
  labs(
    x = "Survival Outcome",
    y = "Beak Length (mm)",
    title = "Mean beak lengths vs. survival"
  )
```


#### Export your figure

```{r}
ggsave("Beak Length Histogram.png", 
       width = 3, height = 3, units = "in")
```

### t-test

Plotting the means and confidence intervals of measurements in the two groups of birds is an important step in data analysis. To perform a more rigourous comparison, however, you may run a t-test to tell if the difference in means is likely due to chance (i.e. sampling error) or not.

The t-test function in R is a bit picky... it requires you to give it two arguments: a vector of measurements for one group and another vector of measurements for the other group. To do this, you must perform three steps:

1. Create a vector of beak lengths for birds that died.
2. Create a vector of beak lengths for birds that survived.
3. Compare the two vectors using a t-test.

STEP 1: The following code will take a dataset, in this case `finches`, remove rows that do not meet a given criteria, and extract the column of interest as a vector, and store the result as a new object named `beak_length_died`.

```{r}
beak_length_died <-
  finches %>% 
  filter(outcome == "died") %>% 
  pull(beak_length)
beak_length_died  # print the new object in the console
```

STEP 2: Now copy that code and change "died" to "survived" in your object name, so `beak_length_survived`, and in the filter criteria, so `outcome == "survived"`.

```{r}
beak_length_survived <-
  finches %>% 
  filter(outcome == "survived") %>% 
  pull(beak_length)
beak_length_survived
```

STEP 3: Perform a t-test using the `t.test()` function:

```{r}
t.test(beak_length_died, beak_length_survived)
```

The two numbers to focus on here are the **p-value** and the **95% confidence interval**.

The *p*-value tells you what proportion of the time you would expect to find a difference in population like these two if they were in fact the same. Scientists generally use a cutoff (or *alpha level*) of 0.05 to determine if a difference is "statistically significant". In this case, a *p*-value of 0.0004539 is much lower than 0.05, so we can reject our null hypothesis and accept our alternative hypothesis that the true difference in means is not equal to 0.

As further evidence that the difference between the means is not equal to zero, you can see that the 95% confidence interval does not include zero.

To report the findings of a t-test, you say:

    Mean beak depths were 10 mm in survivors and 8 mm in non-survivors (Figure 1). This difference was stastistically significant according to a Welch's two-sample t-test assuming unequal variances (t=, df=, p=).

[add link to a SEM explainer]

When you're ready, you can move on to Part 4.

### [Part 4. Bluestem Data Assignment  &rarr;](lab1_4_bluestem.html)