---
title: "Data Lab 2"
---

In this lab you will apply what you learned in [Data Lab 1](lab1.html) to analyze the fast plant data generated by the class this semester.

Remember, the question you are trying to answer is:

<blockquote class="text-success">Does the number of trichomes on Fast Plants (<i>Brassica rapa</i>) plants vary between the F0 (parent) and F1 (offspring) generations.</blockquote>

The results of this assignment will form the RESULTS section of your **Fast Plant Lab Report**.

#### Fast plant data

You can download the fast plant data here:

<a href="fastplant_data.xlsx"><i class="fas fa-file-excel fa-5x"></i></a>

Each row in the dataset represents an individual plant.

The variables in the dataset are:

- **generation**. The generation the plant belonged to? (values include `"F0"` or `"F1"`)
- **trichome_number**. The number of trichomes the plant had. (values are numeric integers)
- **selected**. For the F0 generation, was the individual plant selected for breeding? (values include, `"yes"`, `"no"`, or `NA` if the plant was from the F1 generation)
- **seed_source** For the F1 generation, was the plant from the current year or a previous year (values include `"current_year"`, `"previous_year"`, or `NA` if the plant was from the F0 generation).

#### Your assignment

Your object in this lab is to generate the results that will answer this question, specifically:

- A histogram showing how the distribution of numbers of trichomes varied between generations
- A bar chart showing how the mean number of trichomes varied between generations. Include a 95% confidence interval around the means.
- Figure legends describing these two figures.
- A paragraph describing the results of the analysis, including the results of a *t*-test to test for a difference between the means.

To answer the question, you should:

1. Create a new RStudio project.
2. Copy the code from the [Data Lab 1 beak length script  &rarr;](lab1_beak_length_script.html) [click link] and paste it into a new script.
3. Modify the script to use the fast plant data `"fastplant_data.xlsx"`.
4. Plot a histogram for each generation (F0 and F1) but color the histogram based on whether the plants were selected for breeding or not. For help, see [A new variation] below.

For a reminder on how to perform the analysis steps, see [Data Lab 1, Part 4. Bluestem Data Assignment &rarr;](lab1_4_bluestem.html)

When you are finished, paste your figures into a Word document, add figure legends and a Results paragraph, and upload it to the D2L assignment "Data Lab 2".

#### A new variation

In the finch and bluestem analyses, you created a histogram for each group and also colored the histograms differently for each group.

In theory, there is no reason you can't use a different variable to color the histograms.

For example, here is the finch data colored by `sex`:

```{r read-finch-data, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
source("beak_length.R")
```

```{r finch-sex-histogram, fig.width=6, fig.height=3.5, fig.align='center'}
# histogram of beak length, grouped by survival, colored by sex
ggplot(
  data = finches,
  mapping = aes(x = beak_length,
                fill = sex)           # color the boxes by gender
) +
  geom_histogram(bins = 14) +
  facet_wrap(~ outcome, ncol = 1) +   # put each outcome in its own panel
  labs(
    title = "Distribution of beak lengths",
    subtitle = "Grouped by bird sex and survival outcome",
    x = "Beak Length (mm)",
    y = "Number of Birds"
  )
```

For **your** current assignment, put each generation of plants on its own panel, and color the columns by whether the plants were selected for breeding or not.

For example:

```{r fast-plant-histogram, fig.width=6, fig.height=3.5, fig.align='center'}
read_excel("fastplant_data.xlsx") %>% 
  ggplot(aes(trichome_number)) +
  geom_histogram(aes(fill = selected), bins = 12) +
  facet_wrap(~ generation, ncol = 1) +
  labs(
    title = "Distribution of trichome counts",
    subtitle = "Grouped by generation and selection",
    x = "Number of trichomes",
    y = "Number of plants"
  )
```



