---
title: "Data Lab 2"
---

## Introduction

In this lab you will apply what you learned in [Data Lab 1](lab1.html) to analyze the fast plant data generated by the class this semester.

Follow along with your instructor in class for an introduction to Fast Plants and the question you will be addressing in this lab.

To summarize, the question you are trying to answer is:

<blockquote class="text-success">Does the number of trichomes on Fast Plants (<i>Brassica rapa</i>) plants vary between the F0 (parent) and F1 (offspring) generations.</blockquote>

The figures and results you produce in this **Data Lab 2** are due on D2L in the Assignment folder [Data Lab 2 Fast Plant Results](https://mnstate.learn.minnstate.edu/d2l/le/content/5076831/viewContent/45750038/View). See D2L for the due date. If the link above does not work, you can navigate to the Assignment foldering on D2L and going to Assessment > Assignments > Data Lab 2 Fast Plant Results.

Be sure to consult the [Data Lab 2 Rubric](https://mnstate.learn.minnstate.edu/d2l/le/content/5076831/viewContent/45749771/View) to make sure you have included all necessary components.


This assignment will form the RESULTS section of your **Fast Plant Lab Report**. Your instructor will provide feedback in D2L on your Data Lab 2 assignment submission. You can use that feedback to make changes to your results, which will then form the RESULTS section of your Fast Plant Lab Report. The report will be due on D2L after Data Lab 2 (see D2L for the link and the due date).

## Start your project

Follow these directions to get started:

1. Go to [RStudio Cloud](https://rstudio.cloud/) and sign in. 
2. On the left, under Spaces, click on **Org Bio Workspace**
3. Click the **Start** button on the **Data Lab 2** assignment.

When your project starts, look at the **Files** tab in the lower right pane.

There are two files you should take note of:
- fastplant_data.xlsx - this is the data file you will use
- beak_length.R - this is a copy of the [Beak Length R Script &rarr;](https://biol115l-msum.github.io/lab1_beak_length_script.html) from [Data Lab 1, Part 3. Finch Example &rarr;](https://biol115l-msum.github.io/lab1_3_finch.html).

## Create a new script

We suggest you start your assignment by creating a new script and copying the code from beak_length.R into it. Alternatively you can open **beak_length.R** and go to File > Save As... to save a copy of the script. A good name might be something like **fastplant.R**.

## Look at the fast plant data

In your new script, adjust the code so that it reads the fastplant_data.xlsx instead of finches_data.xlsx. Run the code and note the data object which appeared in your Environment tab. Click on the blue table icon to the right of the object name to open the data viewer window.

Inspect the data.

Each row in the dataset represents an individual plant.

The variables in the dataset are:

- **generation**. The generation the plant belonged to? (values include `"F0"` or `"F1"`)
- **trichome_count**. The number of trichomes the plant had. (values are numeric integers)
- **selected**. For the F0 generation, was the individual plant selected for breeding? (values include, `"yes"`, `"no"`, or `"NA"` if the plant was from the F1 generation)

Ask yourself:

- Which variable contains the numeric variable that was measured or counted? This is the response variable, which you will summarize by calculating the mean.
- Which variable would you use to group the plants? Remember you are trying to compare the response variable at two points in time. Which variable groups plants into two points in time?

## Your assignment

Your objective in this lab is to generate the results that will answer the question above, specifically by producing the following:

- One histogram showing how the distribution of numbers of trichomes varied between generations. It should also show which plants in the F0 generation were selected for breeding and which were not.
- One bar chart showing how the mean number of trichomes varied between generations. Include a 95% confidence interval around the means.
- Two figure legends describing these figures.
- One paragraph describing the results of the analysis, including the results of a *t*-test to test for a difference between the means.

To answer the question, you should:

1. Modify remaining code for use with the fast plant data
2. Plot a histogram with facets for each generation (F0 and F1)
    - Instead of setting the fill color to vary by generation, set it to vary by whether or not the plants were **selected** for breeding. This is the only place you will use the **selected** variable
    - Remove the line `guides(fill = FALSE) +` to show the fill color legend
    - For help, see [A new variation] below
3. Summarize trichome count by generation
4. Create a bar chart showing the mean trichome count in each generation (with confidence intervals)
5. Conduct a t-test to test if the difference in means is statistically significant

For a reminder on how to perform the analysis and writing steps, see [Data Lab 1, Part 4. Bluestem Data Assignment &rarr;](lab1_4_bluestem.html).

When you are finished:

- Paste your two (2) figures into a Word document, add figure legends and a RESULTS paragraph, and upload it to your *lab* D2L page under Assessments > Assignments > Data Lab 2.

### A new variation

In the finch and bluestem analyses, you created a histogram for each group and also colored the histograms differently for each group.

In theory, there is no reason you can't use a different variable to color the histograms.

For example, here is the finch data colored by `sex`:

```{r read-finch-data, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
source("beak_length.R")
file.remove("Beak Length Histogram.png")
file.remove("Beak Length Bar Chart.png")
```

```{r finch-sex-histogram, fig.width=6, fig.height=3.5, fig.align='center'}
# histogram of beak length, grouped by survival, colored by sex
ggplot(
  data = finches,
  mapping = aes(x = beak_length,
                fill = sex)           # color the boxes by gender
) +
  geom_histogram(bins = 14) +
  facet_wrap(~ outcome, ncol = 1) +   # put each outcome in its own panel
  labs(
    title = "Distribution of beak lengths",
    subtitle = "Grouped by sex and outcome",
    x = "Beak Length (mm)",
    y = "Number of Birds"
  )
```

For your current assignment, put each generation of plants on its own panel, and set the fill color of the columns by whether the plants were **selected** for breeding or not.

Remove the line `guides(fill = FALSE) +` to show the fill color legend.

In *your* figures, do not use subtitles and set your titles to say `"Figure 1."` and `"Figure 2."`.

For example:

```{r fast-plant-histogram, fig.width=6, fig.height=3.5, fig.align='center', echo=FALSE}
read_excel("fastplant_data.xlsx") %>% 
  ggplot(aes(trichome_count)) +
  geom_histogram(aes(fill = selected), bins = 12) +
  facet_wrap(~ generation, ncol = 1) +
  scale_fill_discrete(
    name = "Selected for\nbreeding?",
    breaks = c("no", "yes", "NA"),
    labels = c("No", "Yes", "F1")
  ) +
  labs(
    title = "Distribution of trichome counts",
    subtitle = "Grouped by generation and selection status",
    x = "Number of trichomes",
    y = "Number of plants"
  )
```



